RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-11-06 abstract.tex
Small amendmends thoughout the paper.
MASTERFIX

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. Thus, we combine a traditional goal modelling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a methodology, formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model. 
\end{abstract}

>>>>

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. Thus, we combine a traditional goal modelling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a methodology, formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model.

\keywords{Goal Modeling \and Argumentation \and Goal-oriented requirements engineering \and Practical Reasoning} 
\end{abstract}

### 2017-11-06 background.tex
MASTERFIX

In this section, we first introduce our running example, after which we introduce the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, which is the goal modeling language we use to integrate with the argumentation framework. Next, we introduce argumentation: we discuss the \emph{practical reasoning argument scheme (PRAS)}~\cite{atkinson2007}, an argument scheme that is used to form arguments and counter-arguments about situations involving goals, and we give informal examples of how argument and counterargument can influence the status of beliefs about goals. Finally, we briefly discuss the possibilities integrating PRAS and GRL.  %This will be our starting point in the next section.

@@

Our examples and case study are based on the data produced by a recent series of experiments by Schriek et al. \cite{SchriekEtal2016}, who in turn base their work on the so-called Irvine experiment~\cite{UCIworkshop}, which presents a well-known design reasoning assignment in software engineering about a traffic simulator. In this assignment (see Appendix~\ref{sect:designprompt}), designers are provided with a problem description, requirements, and a description of the desired outcomes. The client of the project is Professor E, who teaches civil engineering courses at an American university. In order for the professor to teach students the various theories concerning traffic (such as queuing theory), traffic simulator software needs to be developed in which students can create visual maps of an area, regulate traffic, and so forth. Schriek et al. asked designers (groups of students) to discuss the requirements of this traffic simulator. These discussions were recorded and transcribed, and we used these transcripts for an extensive case study on the basis of which we developed our RationalGRL framework. Furthermore, we also use the traffic simulator case for the running examples throughout this paper.

\subsection{Goal-oriented Requirements Language (GRL)}
\label{sect:background:grl}
GRL is a visual modeling language for specifying intentions, business goals, and non-functional requirements of multiple stakeholders \cite{Amyot:2010:EGM:1841349.1841356}. GRL is part of the User Requirements Notation, an ITU-T standard, that combines goals and non-functional requirements with functional and operational requirements (i.e. use case maps). GRL can be used to specify alternatives that have to be considered, decisions that have been made, and rationales for making decisions. A GRL model is a connected graph of intentional elements that optionally are part of the actors. The GRL elements and relationships used in this paper are shown in Figure~\ref{fig:grl_legend}. Note that we only consider a subset of the full GRL specification in this paper. In particular, the full GRL specification considers different levels of positive and negative contributions and satisfaction levels of goals and tasks, which are not directly relevant for this work. Furthermore, the full GRL specification includes beliefs to capture design rationale, whereas in this paper, the reasoning and rationales behind the models are captured using arguments -- for a brief discussion on beliefs, see Section~\ref{sect:goalmodeling:relatedwork}.

@@

Figure~\ref{fig:example-small} illustrates a simplified GRL diagram from the traffic simulator design exercise. An actor represents a stakeholder of a system or the system itself (\emph{Traffic Simulator}, Figure~\ref{fig:example-small}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals differentiate themselves from goals in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \emph{Realistic simulation}) are often related to non-functional requirements, whereas goals (such as \emph{Generate Cars}) are related to functional requirements. Tasks represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:example-small}, we have the two tasks \emph{Create new cars} and \emph{Keep same cars}: in order to achieve goal \emph{Generate cars}, the simulation can either constantly generate new ones or keep the same cars and have them reappear after they disappear off screen. In order to be achieved or completed, softgoals, goals, and tasks may require resources to be available (e.g., \emph{Car Objects}).

Different links connect the elements in a GRL model. AND, IOR (Inclusive OR), and XOR (eXclusive OR) decomposition links allow an element to be decomposed into sub-elements. In Figure~\ref{fig:example-small}, the goal \emph{Generate cars} is XOR-decomposed to the tasks \emph{Create new cars} and \emph{Keep same cars}, as they are alternative ways of achieving the goal \emph{Generate cars}. Contribution links indicate impacts of one element on another element, which can be positive or negative. Task \emph{Create new cars} has a positive contribution to the task \emph{Realistic simulation}, and a negative contribution to the task \emph{Simple design}. Dependency links model relationships between actors or resources. Here, the goal \emph{Generate cars} depends on the resource \emph{Car objects}.

GRL is based on $i*$~\cite{yu1997towards} and the NFR Framework~\cite{chung2012non}, but it is less restrictive. Intentional elements and links can be more freely combined, the notion of agents is replaced with the more general notion of actors and a task does not necessarily have to be an activity performed by an actor, but may also describe properties of a solution. GRL has a well-defined syntax and semantics. Furthermore, GRL provides support for providing a scalable and consistent representation of multiple views/diagrams of the same goal model (see~\cite[Ch.2]{Ghanavati2013} for more details). GRL is also linked to Use Case Maps, which provide traceability between concepts and instances of the goal model and behavioral design models. Multiple views and traceability links are a good fit with our current research: we aim to add traceability links between intentional elements and their underlying arguments.

GRL has six evaluation algorithms which allow the analysis of alternatives and design decisions by calculating the satisfaction value of the intentional elements quantitatively, qualitatively or in a hybrid way. jUCMNav, GRL tool-support, also allows for adding new GRL evaluation algorithms~\cite{jUCMNav}. GRL also has the capability to be extended through metadata, links, and external OCL constraints. This allows GRL to be used in many domains without the need to change the whole modeling language.

@@

\begin{itemize}
	\item Why is softgoal \emph{Easy to use} not linked to any of the goals or tasks? 
	\item What does \emph{Keep same cars} mean?
	\item Why does task \emph{Create new cars} contribute negatively to \emph{Simple design} and positively to \emph{Realistic simulation}?
	\item Why does \emph{Generate cars} XOR-decompose into two tasks?
\end{itemize}

These are the type of the questions that we cannot answer by just looking at GRL models. The model in Figure~\ref{fig:example-small} does not contain information about discussions that led to the model, such as clarification steps for the naming, or alternatives that have been considered for the relationships. The idea behind the original GRL specification is that beliefs can be used to capture such design rationales that make later justification and review of a model easier. However, beliefs cannot be connected to links - this makes answering the third and fourth question above impossible. Furthermore, beliefs are after-the-fact design rationales and do not capture the types of critical questions given above.

@@

Here, $G$ and $A$ are variables, which can be instantiated with concrete goals and actions to provide a specific practical argument. For example, a concrete argument about the traffic simulator is as follows: 
\begin{itemize}
\item[] \emph{Generate Cars} is a goal,
\item[] Performing action \emph{Keep same cars} realizes goal \emph{Generate cars}, 
\item[] \underline{Therefore} 
\item[] Action \emph{Keep same cars} should be performed
\end{itemize}

Note that PRAS is an argument scheme that captures a full inference step: ``$G$, $A$ realizes $G$ \emph{Therefore} $A$''. There are, however, also schemes that capture simpler reasoning patterns, such as claims of the form ``$A$ does not realize $G$''. We will discuss these schemes below.

In argumentation, conclusions which are at one point acceptable can later be rejected because of new information. For example, we may argue that, in fact, performing action \emph{Keep same cars} does not realize goal \emph{Generate cars}, thus giving a counterargument to the above instantiation of PRAS. Atkinson et al.~\cite{atkinson2007} define a set of so-called critical questions that point to typical ways in which an argument based on PRAS can be criticized by. Some examples of critical questions are as follows.

@@

The idea is that answers to critical questions are counterarguments to the original PRAS argument. These counterarguments also follow a scheme; for example, a negative answer to CQ1 follows the scheme ``Action $A$ will not realize goal $G$'', which can be instantiated (e.g. ``\emph{Keep same cars} does not realize \emph{Generate cars''}) to form a counterargument to the original argument.

Another way to criticize an argument for an action is to suggest an alternative action that realizes the same goal (CQ2). For example, we can argue that performing \emph{Create new cars} also realizes the goal \emph{Generate cars}. Also, it is possible that performing an action has a negative side effect (CQ3). For example, while the action \emph{Create new cars} realizes the goal \emph{Generate cars}, it has a negative side effect, namely hurting \emph{Simple design}: having the simulation constantly create new cars is fairly complex design choice.

@@

In argumentation, counterarguments are said to \emph{attack} the original arguments. Given a set of arguments and attacks between these arguments, we can compute which arguments are accepted and which are rejected using different argumentation semantics~\cite{Dung1995}\footnote{Formal definitions of argumentation frameworks and semantics will be given in section \ref{sect:gmas}. In this section, we will briefly discuss the intuitions behind these concepts.}. Figure \ref{fig:pras:example} shows three arguments from the traffic simulation example, where arguments are rendered as boxes and attack relations as arrows. There are two arguments based on PRAS: argument A1 for \emph{Keep Same Cars} and argument A2 for \emph{Create new cars}. Argument A2 proposes an alternative way of realizing the same goal \emph{Generate cars} with respect to argument A1 and vice versa (cf. CQ2), so A1 and A2 mutually attack each other, denoted by the arrows between A1 and A2. Argument A3 says that \emph{Create new cars} has a negative effect on \emph{Static Simulation}, so A3 attacks A2, as it points to a negative side-effect of \emph{Create new cars} (CQ3). The intuition here is that an argument is acceptable if any argument that attacks it is itself rejected. In Figure~\ref{fig:pras:example}, argument A3 is accepted because it has no attackers. This makes A2 rejected (indicated by the lighter grey color), because its attacker A3 is accepted. A1 is then also accepted, since its only attacker, A2, is rejected.

Looking at PRAS and its critical questions, one can see how it could be used to argue about goals and actions or, more specifically, about goal models. However, we cannot literally use PRAS and its critical questions, as there are elements of the GRL language, such as actors and resources, which cannot be found in PRAS. Furthermore, it is not directly clear whether the critical questions as proposed by Atkinson and Bench-Capon~\cite{atkinson2007} actually apply to GRL models. In fact, our case study (Section \ref{sect:gmas}) shows that when discussing requirements, people very often do not structure their reasoning nicely in the way that PRAS presents it. That is, you do not see the discussants setting up an argument ``We have goal $G$, $A$ realizes $G$ \emph{Therefore} we should perform $A$''. A typical discussion is much more unstructured, as is clear from the transcript excerpts in Appendix~\ref{sect:transcripts:excerpts}. So if we would use the version of PRAS presented in this section for our argumentation, we would violate requirement 1: The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase. Our solution was to develop our own set of argument schemes and critical questions by analyzing transcripts of discussions about the traffic simulator. This set of schemes and questions and our case study are described in the next section.

>>>>

In this section, we first introduce our running example, after which we introduce the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, which is the goal modeling language we use to integrate with the argumentation framework. In this article, we consider a slightly simplified version of GRL. We motivate this and explain which subset of GRL we consider. Next, we introduce argumentation: we discuss the \emph{practical reasoning argument scheme (PRAS)}~\cite{atkinson2007}, an argument scheme that is used to form arguments and counter-arguments about situations involving goals, and we give informal examples of how argument and counterargument can influence the status of beliefs about goals. Finally, we briefly discuss the possibilities integrating PRAS and GRL.  %This will be our starting point in the next section.

@@

Our examples and case study are based on the data produced by a recent series of experiments by Schriek et al. \cite{SchriekEtal2016}, who in turn base their work on the so-called Irvine experiment~\cite{UCIworkshop}. This experiment contains a well-known design reasoning assignment in software engineering about a traffic simulator. In this assignment (see Appendix~\ref{sect:designprompt}), designers are provided with a problem description, requirements, and a description of the desired outcomes. The client of the project is Professor E, who teaches civil engineering courses at an American university. In order for the professor to teach students the various theories concerning traffic (such as queuing theory), traffic simulator software needs to be developed in which students can create visual maps of an area, regulate traffic, and so forth. Schriek et al. asked designers (groups of students) to discuss the requirements of this traffic simulator. These discussions were recorded and transcribed, and we used these transcripts for an extensive case study on the basis of which we develop our RationalGRL framework. Furthermore, we also use the traffic simulator case for the running examples throughout this paper.

\subsection{Goal-oriented Requirements Language (GRL)}
\label{sect:background:grl}
GRL is a visual modeling language for specifying intentions, business goals, and non-functional requirements of multiple stakeholders \cite{Amyot:2010:EGM:1841349.1841356}. GRL is part of the User Requirements Notation, an ITU-T standard, that combines goals and non-functional requirements with functional and operational requirements (i.e. use case maps). GRL can be used to specify alternatives that have to be considered, decisions that have been made, and rationales for making decisions. A GRL model is a connected graph of intentional elements that optionally are part of the actors. The GRL elements and relationships used in this paper are shown in Figure~\ref{fig:grl_legend}.

@@

Figure~\ref{fig:example-small} illustrates a simplified GRL diagram from the traffic simulator design exercise. An actor represents a stakeholder of a system or the system itself (\emph{Traffic Simulator}, Figure~\ref{fig:example-small}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals differentiate themselves from goals in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \emph{Realistic simulation}) are often related to non-functional requirements, whereas goals (such as \emph{Generate Cars}) are related to functional requirements. Tasks represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:example-small}, there are two tasks \emph{Create new cars} and \emph{Keep same cars}: in order to achieve goal \emph{Generate traffic}, the simulation can either constantly generate new ones or keep the same cars and have them reappear after they disappear off screen. In order to be achieved or completed, softgoals, goals, and tasks may require resources to be available (e.g., \emph{Car Objects}).

Different links connect the elements in a GRL model. AND, IOR (Inclusive OR), and XOR (eXclusive OR) decomposition links allow an element to be decomposed into sub-elements. In Figure~\ref{fig:example-small}, the goal \emph{Generate traffic} is XOR-decomposed to the tasks \emph{Create new cars} and \emph{Keep same cars}, as they are alternative ways of achieving the goal \emph{Generate traffic}. Contribution links indicate impacts of one element on another element, which can be positive or negative. Task \emph{Create new cars} has a positive contribution to the softgoal \emph{Realistic simulation}, and a negative contribution to the softgoal \emph{Simple design}. Dependency links model relationships between actors or resources. Here, the goal \emph{Generate traffic} depends on the resource \emph{Car objects}.

GRL is based on $i*$~\cite{yu1997towards} and the NFR Framework~\cite{chung2012non}, but it is less restrictive. Intentional elements and links can be more freely combined, the notion of agents is replaced with the more general notion of actors and a task does not necessarily have to be an activity performed by an actor, but may also describe properties of a solution. GRL has a well-defined syntax and semantics. Furthermore, GRL provides support for providing a scalable and consistent representation of multiple views/diagrams of the same goal model (see~\cite[Ch.2]{Ghanavati2013} for more details). GRL is also linked to Use Case Maps, which provide traceability between concepts and instances of the goal model and behavioral design models. Multiple views and traceability links are a good fit with our current research: we aim to add traceability links between intentional elements and their underlying arguments. GRL has six evaluation algorithms which allow the analysis of alternatives and design decisions by calculating the satisfaction value of the intentional elements quantitatively, qualitatively or in a hybrid way.

GRL is implemented in the open-source tool called jUCMNav~\cite{jUCMNav}, which is actively maintained\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}. jUCMNav currently has over 2.000 commits made by over 40 contributors, representing over 200.000 lines of code. jUCMNav is used actively in research, it is mentioned in over 400 papers according to Google Scholar. As jUCMNav is open, other research teams have also extended the tool in many directions in the past few years, e.g., for: modeling and analyzing security misuse cases, supporting activity theory, Web-based goal modeling, and enterprise architecture modeling. jUCMNav also allows for adding new GRL evaluation algorithms. GRL also has the capability to be extended through metadata, links, and external OCL constraints. This allows GRL to be used in many domains without the need to change the modeling language itself.

@@

\begin{itemize}
	\item Why is softgoal \emph{Easy to use} not linked to any of the goals or tasks? 
	\item What does \emph{Keep same cars} mean?
	\item Why does task \emph{Create new cars} contribute negatively to \emph{Simple design} and positively to \emph{Realistic simulation}?
	\item Why does \emph{Generate traffic} XOR-decompose into two tasks?
\end{itemize}

These are the type of the questions that we cannot answer by just looking at GRL models. The model in Figure~\ref{fig:example-small} does not contain information about discussions that led to the model, such as clarification steps for the naming, or alternatives that have been considered for the relationships.

In this article, we consider a subset of the full GRL specification. In particular, the full GRL specification considers different levels of positive and negative contributions and satisfaction levels of goals and tasks, which are not directly relevant for this work. Therefore, we consider only two kinds of contribution values in our framework: \emph{positive} and \emph{negative}. It is straightforward to extend this to multiple values, but we have not done so to keep our presentation concise.

Furthermore, the full GRL specification includes beliefs to capture design rationale, whereas in this paper, the reasoning and rationales behind the models are captured using arguments. The idea behind the original GRL specification is that beliefs can be used to capture such design rationales that make later justification and review of a model easier. While beliefs can indeed be used for rationalization by connecting them to IEs, jUCMNav does not use them in the evaluation algorithms. Moreover, a belief can only be used to \emph{support} IEs, not to reject them. For a more detailed comparison of our framework with GRL beliefs, see Section~\ref{sect:goalmodeling:relatedwork}.

@@

Here, $G$ and $A$ are variables, which can be instantiated with concrete goals and actions to provide a specific practical argument. For example, a concrete argument about the traffic simulator is as follows: 
\begin{itemize}
\item[] \emph{Generate traffic} is a goal,
\item[] Performing action \emph{Keep same cars} realizes goal \emph{Generate traffic}, 
\item[] \underline{Therefore} 
\item[] Action \emph{Keep same cars} should be performed
\end{itemize}

Note that PRAS is an argument scheme that captures a full inference step: ``$G$, $A$ realizes $G$, \emph{Therefore} $A$''. There are, however, also schemes that capture simpler reasoning patterns, such as claims of the form ``$A$ does not realize $G$''. We will discuss these schemes below.

In argumentation, conclusions which are at one point acceptable can later be rejected because of new information. For example, we may argue that, in fact, performing action \emph{Keep same cars} does not realize goal \emph{Generate traffic}, thus giving a counterargument to the above instantiation of PRAS. Atkinson et al.~\cite{atkinson2007} define a set of so-called critical questions that point to typical ways in which an argument based on PRAS can be criticized. Some examples of critical questions are as follows.

@@

The idea is that answers to critical questions are counterarguments to the original PRAS argument. These counterarguments also follow a scheme; for example, a negative answer to CQ1 follows the scheme ``Action $A$ will not realize goal $G$'', which can be instantiated (e.g. ``\emph{Keep same cars} does not realize \emph{Generate traffic''}) to form a counterargument to the original argument.

Another way to criticize an argument for an action is to suggest an alternative action that realizes the same goal (CQ2). For example, we can argue that performing \emph{Create new cars} also realizes the goal \emph{Generate traffic}. Also, it is possible that performing an action has a negative side effect (CQ3). For example, while the action \emph{Create new cars} realizes the goal \emph{Generate traffic}, it has a negative side effect, namely hurting \emph{Simple design}: having the simulation constantly create new cars is fairly complex design choice.

@@

In argumentation, counterarguments are said to \emph{attack} the original arguments. Given a set of arguments and attacks between these arguments, we can compute which arguments are accepted and which are rejected using different argumentation semantics~\cite{Dung1995}\footnote{Formal definitions of argumentation frameworks and semantics will be given in Section~\ref{sect:gmas}. In this section, we briefly discuss the intuitions behind these concepts.}. Figure \ref{fig:pras:example} shows three arguments from the traffic simulation example, where arguments are rendered as boxes and attack relations as arrows. There are two arguments based on PRAS: argument A1 for \emph{Keep Same Cars} and argument A2 for \emph{Create new cars}. Argument A2 proposes an alternative way of realizing the same goal \emph{Generate traffic} with respect to argument A1 and vice versa (cf. CQ2), so A1 and A2 mutually attack each other, denoted by the arrows between A1 and A2. Argument A3 says that \emph{Create new cars} has a negative effect on \emph{Static Simulation}, so A3 attacks A2, as it points to a negative side-effect of \emph{Create new cars} (CQ3). The intuition here is that an argument is acceptable if any argument that attacks it is itself rejected. In Figure~\ref{fig:pras:example}, argument A3 is accepted because it has no attackers. This makes A2 rejected (indicated by the lighter grey color), because its attacker A3 is accepted. A1 is then also accepted, since its only attacker, A2, is rejected.

Looking at PRAS and its critical questions, one can see how it could be used to argue about goals and actions or, more specifically, about goal models. However, we cannot literally use PRAS and its critical questions, as there are elements of the GRL language, such as actors and resources, which cannot be found in PRAS. Furthermore, it is not directly clear whether the critical questions as proposed by Atkinson and Bench-Capon~\cite{atkinson2007} actually apply to GRL models. In fact, our case study (Section \ref{sect:gmas}) shows that when discussing requirements, people very often do not structure their reasoning nicely in the way that PRAS presents it. That is, you do not see the discussants setting up an argument ``We have goal $G$, $A$ realizes $G$ \emph{Therefore} we should perform $A$''. A typical discussion is much more unstructured, as is clear from the transcript excerpts in Appendix~\ref{sect:transcripts:excerpts}. So if we would use the version of PRAS presented in this section for our argumentation, we would violate requirement 1: The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase. Our solution is to develop our own set of argument schemes and critical questions by analyzing transcripts of discussions about the traffic simulator. This set of schemes and questions and our case study are described in the next section.
